---
title: "SUARE Bericht_v40 Bildung"
output: html_document
date: "2024-08-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Pre-work

Load packages

```{r}
library(pacman)

pacman::p_load(tidyverse,
               haven, #read .dta dta
               naniar,
               
               sjmisc, # search for variable labels in dataset
               sjPlot) #print frq tables


```


## 1.1. Sample and Data cleaning

```{r}
data_v40 <- read_dta("") # forer version of data-file

```


```{r}
# Exploring: How many unique HH and pids are in the dataset?

data_v40 |>
  summarise(n_distinct(hid),
            n_distinct(pid)) # 2219 hids and 3403 pids


# generate basic socio-demographic variabes
data_v40 <- data_v40 |>
  mutate(age = syear - geburt, # age
         sex_bin = case_when( # gender (binary)
           sex == 1 ~ 0, # male
           sex == 2 ~ 1, # female
           TRUE ~ NA
         )) |>
  set_variable_labels(sex_bin = "Gender") |>
  set_value_labels(sex_bin = c("male" = 0, "female" = 1))
  

# exploring: variables in dataset indicating if pids and hids have/dont have children and how many
data_v40 |>
  frq(lr3192) # [Children] yes= 2,590; no= 792; missings= 21

data_v40 |>
  frq(lb0285) # 1-more = 2,579; missings = 824

data_v40 |>
  frq(prev_hlk0044_v2) # [Children (<16 y.o.) in HH] yes= 1,812; no= 1,591

data_v40 |>
  frq(prev_nrkid) # [how many children (<16) in HH] 0=1,591, 1-more=1,812


# isolating only his/pids with children (<16) in HH
data_v40 <- data_v40 |>
  filter(prev_hlk0044_v2 == 1)


# how many unique pids/hids are in the dataset (having children < 16 in HH)
data_v40 |>
  summarise(n_distinct(hid), # 1812
            n_distinct(pid)) # 1122 


# how many persons are in HHs, considering only HH with children <16
## Generate variable that counts how many pid by hid are present
data_v40 <- data_v40 |>
  group_by(hid) |>
  mutate(n_pid_HH = n()) |>
  ungroup()

## Frequency table of the n_pid_HH variable
data_v40 |>
  frq(n_pid_HH)

## set variable label: 
data_v40 <- data_v40 |>
  set_variable_labels(n_pid_HH = "Total number of persons in household")


# how many unique hids and pids are in dataset considering the number of persons in househild
summary <- data_v40 |>
  group_by(n_pid_HH) |>
  summarise(
    total_hid = n(),
    distinct_hid = n_distinct(hid),
    total_pid = n(),
    distinct_pid = n_distinct(pid)
  ) |>
  arrange(n_pid_HH)

print(summary)

```

## 1.2. Cleaned Dataset

Load data

```{r}
data_v40c <- read_dta("I:/MA/amarchitto/SUARE/Bericht_finale_dataset/temp/suare_bericht_v40_Bildung_cleaned.dta")
```


THEMEN:

1) Kinderbetreuungsbedarfe und Prävalenz
2) Verteilung Kinder Schultypen (Schulsystem/Bundeslandgruppe)
3) Weiterführende Bildung und Bildungsbeteiligung: Ausbildung / Uni


VERWENDETE ITEMS:

HH-Fragebogen (Betreeuungsbatterie / Schultyp)
Personenfragebogen (Weiterführende Bildung, Soziodemoghraphische Infos über Eltern)




Inspect data

```{r}

#Search for variable labels 
find_var(data_v40c,
         pattern = c("prev_k_birthy_v2_"),
         search = c("all"),
         fuzzy = TRUE) 


# Show variable values
frq(data_v40c, lr3192)
```

VARIABLEN:

- pid, hid,
- syear
- lb0288_v2: Kind: Geschlecht (1)	
- ks_pre_v6_: Kinder (<16) im HH: Besuch Betreuungseinrichtung (1)	
- kc_kindbetr_1: Kinder (<16) im HH: Besuch Betreuungseinrichtung (Std./Woche) (1)
- lb0287_v2: Kind: Geburtsjahr (1)
- lkvpid1: Kind: Unveränderliche Personennummer (1)
- hknr_1: Personennummer Kind (1)
- ks_gen_v8_1: Kinder (<16) im HH: Schule (1)
- ks_stufe_1: Kinder (<16) im HH: Klassenstufe (1)
- ks_asc_v2_1: Kinder (<16) im HH: Hortbesuch (1)


### 1.2.1. Clean Cleaned Dataset


```{r}
# Select variables that I need
data_v40c_cleaned <- data_v40c |>
  select(1:10, 
         pid, age, sex_bin,syear,
         n_pid_HH, # Anzahl Personen in HH 
         lr3192, prev_nrkid, #Kinder
         hknr_1:hknr_10, # Personennummer Kind
         starts_with("ks_pre_v6_"), # Betreuungseinrichtung
         #starts_with("kc_kindbetr_"), # Betreuungseinrichtung Stunden/Woche
         starts_with("ks_gen_v8_"), # Schule
         starts_with("prev_k_birthy_v2_"), 
         #starts_with("ks_stufe_"), # Klassenstufe
         #starts_with("ks_asc_v2_"), # Hortbesuch
         #lb0288_v2, lb0291_v1, lb0294_v1, lb0297_v1, lb0300_v1, lb0303_v1	, lb0306_v1, lb0309_v1, lb1166_v1, lb1165, # Geschlecht Kind
         lb0287_v2, lb0290_v2, lb0293_v2, lb0296_v2,lb0299_v2, lb0302_v2,  lb0305_v2, lb0308_v2, lb1139, lb1138, # Geburtsjahr Kind
         )


# rename Geburtsjahr variable kid
data_v40c_cleaned <- data_v40c_cleaned |>
  rename(kidborn_1 = lb0287_v2,
         kidborn_2 = lb0290_v2,
         kidborn_3 = lb0293_v2,
         kidborn_4 = lb0296_v2,
         kidborn_5 = lb0299_v2,
         kidborn_6 = lb0302_v2,
         kidborn_7 = lb0305_v2,
         kidborn_8 = lb0308_v2,
         kidborn_9 = lb1139,
         kidborn_10 = lb1138)



# Replace missing values with NA:

na_numbers <- c(-8:-1) #this are the missing values that I want NA

data_v40c_cleaned <- data_v40c_cleaned |>
  replace_with_na_all(condition = ~.x %in% na_numbers)

```


# 2. Analysis (Cleaned Dataset)


```{r}
# what is the gender situation of households
data_v40c_cleaned |>
  frq(sex_bin)

# how many children (<16 in HH) are present by hid? 
data_v40c_cleaned |>
  distinct(hid, .keep_all = TRUE) |>
  frq(prev_nrkid)


# how old are children (<16 in HH): exploring vars prev_k_birthy_v2_
test <- data_v40c_cleaned |>
  mutate(across(starts_with("prev_k_birthy_v2_"), ~ 2023 - ., .names = "age_k_{.col}"))

combined_ages <- test |>
  distinct(hid, .keep_all = TRUE) |>
  select(starts_with("age_k_prev_k_birthy_v2_")) |>
  unlist()

table(combined_ages)



# Categorize children into eligibility groups (0-5 for childcare, 6-17 for school)
test_2 <- test |>
  mutate(across(starts_with("age_k_"), ~ case_when(
    . < 6 ~ 1,               # Age 0-5: eligible for childcare
    . >= 6 & . <= 17 ~ 2,   # Age 6-17: eligible for school
    TRUE ~ NA_real_         # Handle any other values (optional)
  ), .names = "age_k_categ_{.col}"))
  



# starting analysis of single parents (pid==hid==539)
single_parents <- test_2 |>
  filter(n_pid_HH == 1) |>
  distinct(hid, .keep_all = TRUE) |>
  select(starts_with("age_k_categ_")) |>
  unlist()
  
table(single_parents)



# how are school types distributed?

school <- test |>
  distinct(hid, .keep_all = TRUE) |>
  select(starts_with("ks_gen_v8_")) |>
  unlist()

table(school)






```








Calculating frequency tables for 1. Child, 2. Child... will be possible however not very effective as I want descriptive statistics for every child


```{r}
# School type
data_v40c_cleaned |>
  frq(ks_gen_v8_1)

# Besuch Betreuungseinrichtung
data_v40c_cleaned |>
  frq(ks_pre_v6_1)

```



For real analysis I first need do some tidying: transform from wide to long format

```{r}
data_v40c_long <- data_v40c_cleaned |>
  pivot_longer(
    cols = starts_with("hknr_"),
    names_to = "kind",
    values_to = "pidkind") |>
  
  pivot_longer(
    cols = starts_with("ks_gen_v8_"),
    names_to = "schulbildung",
    values_to = "schule"
  ) |>
  pivot_longer(
    cols = starts_with("ks_pre_v6_"),
    names_to = "betreuungseinrichtungsbesuch",
    values_to = "betreuungseinrichtung"
  ) |>
  pivot_longer(
    cols = starts_with("kidborn_"),
    names_to = "geburt",
    values_to = "geburtsjahr"
  ) |>
  
  filter(str_sub(kind, -1) == str_sub(schulbildung, -1) & 
           str_sub(kind, -1) == str_sub(betreuungseinrichtungsbesuch, -1) &
           str_sub(kind, -1) == str_sub(geburt, -1)) |>  # filter rows so that only those stay in which the index of the variable names match (z.b. hknr_1 matches ks_gen_v8_1 and ks_pre_v6_1 and ks_asc_v2_1)
  select(-kind, -schulbildung, - betreuungseinrichtungsbesuch, -geburt) # only needed for filtering so I can remove them now
  


```



Clean it further

```{r}
# filter for those observations with children -> 
# Currently there are 10 x 1812 observations as the child variables range from _1 to _10 while most household heads only have a few if any children
data_v40c_long_clean <- data_v40c_long |>
  filter(if_all(c("pidkind"), ~!is.na(.))) 


length(unique(data_v40c_long_clean$pid)) # Descriptive: As there are now duplicates of houshold heads if they have more than one child this shows how many unique pids there are -> 1812 (parents)


# calculate age of children
data_v40c_long_clean <- data_v40c_long_clean |>
  mutate(agekid = syear - geburtsjahr)

# inspect age distribution of kids
data_v40c_long_clean |>
  filter(agekid <= 16) |>
  frq(agekid)




```


## 2.1. Analysis

Schulbesuch

```{r}

analysis_schulsex <- data_v40c_long_clean |>
  select(schule, sex_bin) |>
  mutate(schule = as_factor(schule),
         sex_bin = as_factor(sex_bin)) 


CrossTable(test$schule, test$sex_bin,
           prop.r = TRUE, prop.c = FALSE, prop.t = FALSE, prop.chisq = FALSE)




```


Betreuungsbedarf

```{r}
# Distribution of children age

data_v40c_long_clean |>
  filter(agekid <= 16) |>
  frq(agekid)

```









